### 2021-06-08 papers/thesis/full/ch3-disk-slices/disk-slices.tex
Addressed most comments so far.

In recent years, there has been some interesting work in the area of multivariate orthogonal polynomials in two-dimensional domains (see \cite{dunkl2014orthogonal}), such as for a Chebychev-based spectral method on the rectangle \cite{julien2009efficient}. In particular, spectral methods have been established on the disk \cite{boyd2011comparing, vasil2016tensor} and the triangle \cite{olver2019triangle}. Examples of applications for the disk include use in optics \cite{mahajan2007orthonormal}, astrophysics \cite{pringle1981accretion}, and fluid dynamics \cite{eggels1994fully, noll1976zernike, kerswell2005recent}, and for the triangle include solving Volterra integral equations \cite{gutleb2020sparse}.

For the disk, a review of spectral methods on the disk for solving the Poisson equation has been completed \cite{boyd2011comparing}, that looks at methods involving various bases for expanding functions in the disk including Zernike polynomials and Chebyshev polynomials. It is concluded that using Chebyshev polynomials yields mixed results, but that there is also no single advised basis to use. While Zernike polynomials yield accurate results, there is a lack of radial fast transform, which is a disadvantage. However, ideas for how to overcome this have progressed \cite{slevinsky2018use, sakai2009application, wilber2017computing}. One solution proposed is pairing the Zernike disk OPs with a radial basis defined in terms of Jacobi polynomials \cite{vasil2016tensor}:
\bseqnnumber{
	Q_n^{k,m}(r) := C_n^{k,m} r^m P^{(k,m)}_n(2r^2 - 1), \label{eqn:ds:diskops}.
}
where $C_n^{k,m}$ is some normalising constant so that they are orthonormal with respect to $\int_0^1 Q_n^{k,m}(r) \; Q_{n'}^{k,m}(r) \; (1-r^2)^k \; r \; \D r = \delta_{n,n'}$. $\{Q_n^{k,m}\}$ are therefore a two-parameter family of orthogonal polynomials. The use of polar coordinates here then allows for radially symmetric partial differential (and other) operators, as well as basis (parameter) conversion operators, to be represented as banded matrices by exploiting the relations of the Jacobi polynomials -- they are extremely useful for solving a number of problems in not just 1D but 2D also, since one can utilise the sparsity of the relations that they hold to yield operator matrices that are efficient to solve \cite{doha2006efficient}.

@@

We will show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations on the interval \cite{olver2013fast}, and its analogues on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}. On the disk-slice the family of weights we consider are of the form
\bseqn{
	\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c
}
for $\alpha \leq x \leq \beta,$ $-\rho(x) \leq y \leq \rho(x)$. The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
}
where $C_{n,k} \neq 0$ and \enquote{lower order terms'} includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs we will see in \bsrefdef{def:ds:OPconstruction}.

@@

@@

Further, note that $J_x^{(a,b,c,d)}, J_y^{(a,b,c,d)}$ are banded-block-banded matrices (see \bsrefdef{def:sh:bandedblockbanded}). They are block-tridiagonal (block-bandwidths $(1,1)$):
\bseqn{
J_{x/y}^{(a,b,c,d)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}.
}
For $J_x^{(a,b,c,d)}$, the blocks are diagonal (sub-block-bandwidths $(0,0)$) for $n = 0,1,2,\dots$:
\bseqn{
	A^x_n &:= 
		\begin{pmatrix}
			\alphaabcd_{n+1,0,1} & 0 & \hdots & 0 \\
			& \ddots & & \vdots & \\
			& & \alphaabcd_{n+1,n,1} & 0 \\
		\end{pmatrix} \in \R^{(n+1)\times(n+2)}, \\
	B^x_n &:= 
		\begin{pmatrix}
			\alphaabcd_{n,0,2} & & \\
			& \ddots & \\
			& & \alphaabcd_{n,n,2} \\
		\end{pmatrix} \in \R^{(n+1)\times(n+1)}, \\
	C^x_n &:= \big( A^x_{n-1} \big)^\top \in \R^{(n+1)\times n}, \quad (n \ne 0)
}
and for $J_y^{(a,b,c,d)}$ the blocks are tridiagonal (sub-block-bandwidths $(1,1)$):
\bseqn{
	A^y_n &:= 
		\begin{pmatrix}
			\betaabcd_{n,0,8} & \betaabcd_{n,0,9} & & & \\
			\betaabcd_{n,1,7}& \ddots & \ddots & & \\
			& \ddots & \ddots & \ddots & \\
			& & \betaabcd_{n,n,7} & \betaabcd_{n,n,8} & \betaabcd_{n,n,9} \\
		\end{pmatrix} \in \R^{(n+1)\times(n+2)}, \\
	B^y_n &:= 
		\begin{pmatrix}
			\betaabcd_{n,0,5} & \betaabcd_{n,0,6} & & \\
			\betaabcd_{n,1,4} & \ddots & \ddots & \\
			& \ddots & \ddots & \betaabcd_{n,n-1,6} \\
			& & \betaabcd_{n,n,4} & \betaabcd_{n,n,5}
		\end{pmatrix} \in \R^{(n+1)\times(n+1)}, \\
	C^y_n &:= 
		\begin{pmatrix}
			\betaabcd_{n,0,2} & \betaabcd_{n,0,3} & & \\
			\betaabcd_{n,1,1} & \ddots & \ddots & \\
			& \ddots & \ddots & \betaabcd_{n,n-2,3} \\
			& & \ddots & \betaabcd_{n,n-1,2} \\
			& & & \betaabcd_{n,n,1} \\
		\end{pmatrix} \in \R^{(n+1)\times n}, \quad (n \ne 0).
}
Note that the sparsity of the Jacobi matrices (in particular the sparsity of the sub-blocks) comes from the natural sparsity of the three-term recurrences of the 1D OPs, meaning that the sparsity is not limited to the specific disk-slice case.

@@

In this section, we concentrate on the disk-slice case, and simply note that similar arguments apply for the trapezium case. Recall that, for the disk-slice,
\bseqn{
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
}
where
\bseqn{
	\begin{cases}
		(\alpha,\beta) &\subset (0,1) \\
		(\gamma,\delta) &:= (-1,1) \\
		\rho(x) &:= (1-x^2)^{\half}
	\end{cases}.
}
The 2D OPs on the disk-slice $\Omega$, orthogonal with respect to the weight
\bseqn{
	\Wabc(x,y) &:= \genjacw^{(a,b,2c)}(x) \: \jacw^{(c)}\fpr(\frac{y}{\rho(x)}) \\
	&= (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c, \quad (x,y) \in \Omega,
}
are then given by:
\bseqn{
	\hdopnkabc(x,y) := \genjacnmk^{(a, b, 2c+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(c,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega
}
where the 1D OPs $\{\genjac^{(a, b, c)}_n\}$ are orthogonal on the interval $(\alpha, \beta)$ with respect to the weight
\bseqn{
	\genjacw^{(a,b,c)}(x) :=  (\beta - x)^a \: (x - \alpha)^b \: \rho(x)^{c}
}
and the 1D OPs $\{\jac^{(c, c)}_n\}$ are orthogonal on the interval $(\gamma, \delta) = (-1, 1)$ with respect to the weight
\bseqn{
	\jacw^{(c)}(x) := (1-x)^c \: (1+x)^c = (1-x^2)^c.
}
Denote the weighted OPs by
\bseqn{
	\bigWabc(x,y) := \Wabc(x,y) \: \bighdopabc(x,y),
}
and recall that a function $f(x,y)$ defined on $\Omega$ is approximated by its expansion $f(x,y) = \bighdopabc(x,y)^\top \vec{f}$.

@@

The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomial families' derivatives, for example the Jacobi polynomials on the interval, as seen in the DLMF \cite[(18.9.3)]{DLMF}, and on the triangle \cite{olver2018recurrence}. An illustration of how the non-weighted differential operators increment the parameters (a,b,c) is seen in \bsreffig{fig:ds:laplaceparameterconversion}.

>>>>

In recent years, there has been some interesting work in the area of multivariate orthogonal polynomials in two-dimensional domains (see \cite{dunkl2014orthogonal}), such as for a Chebyshev-based spectral method on the rectangle \cite{julien2009efficient}. In particular, spectral methods have been established on the disk \cite{boyd2011comparing, vasil2016tensor} and the triangle \cite{olver2019triangle}. Examples of applications for the disk include use in optics \cite{mahajan2007orthonormal}, astrophysics \cite{pringle1981accretion}, and fluid dynamics \cite{eggels1994fully, noll1976zernike, kerswell2005recent}, and for the triangle include solving Volterra integral equations \cite{gutleb2020sparse}.

For the disk, a review of spectral methods on the disk for solving the Poisson equation has been completed \cite{boyd2011comparing}, that looks at methods involving various bases for expanding functions in the disk including Zernike polynomials and Chebyshev polynomials. It is concluded that using Chebyshev polynomials yields mixed results, but that there is also no single advised basis to use. Zernike polynomials were found to yield accurate results, and ideas for how to implement a radial transform have progressed in recent years \cite{slevinsky2018use, sakai2009application, wilber2017computing} \bstodo{not sure what to say here?}. One solution proposed is pairing the Zernike disk OPs with a radial basis defined in terms of Jacobi polynomials \cite{vasil2016tensor}:
\bseqnnumber{
	Q_n^{k,m}(r) := C_n^{k,m} r^m P^{(k,m)}_n(2r^2 - 1), \label{eqn:ds:diskops}
}
where $C_n^{k,m}$ is some normalising constant so that they are orthonormal with respect to $\int_0^1 Q_n^{k,m}(r) \; Q_{n'}^{k,m}(r) \; (1-r^2)^k \; r \; \D r = \delta_{n,n'}$. $\{Q_n^{k,m}\}$ are therefore a two-parameter family of orthogonal polynomials. The use of polar coordinates here then allows for radially symmetric partial differential (and other) operators, as well as basis (parameter) conversion operators, to be represented as banded matrices by exploiting the relations of the Jacobi polynomials -- they are extremely useful for solving a number of problems in not just 1D but 2D also, since one can utilise the sparsity of the relations that they hold to yield operator matrices that are efficient to solve \cite{doha2006efficient}.

@@

We will show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations on the interval \cite{olver2013fast}, and its analogues on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}. On the disk-slice the family of weights we consider are of the form
\bseqn{
	\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c
}
for $\alpha \leq x \leq \beta,$ $-\rho(x) \leq y \leq \rho(x)$. The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
}
where $C_{n,k} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs we will see in \bsrefdef{def:ds:OPconstruction}.

@@

\remark We should note at this stage that hereon we will continue to use the notation $\jac_n^{(a, b)}$ to refer to the univariate OPs orthonormal with respect to the weight $\jacw^{(a,b)}$. For the disk-slice case, this means they are the orthonormal Jacobi polynomials, which is why we choose this notation. We stress that this is in difference to the spherical harmonic framework in \bsrefchapter{CHAPTER:sphericalharmonics}, where conventionally they are defined using the un-normalized (i.e. simply orthogonal) classical Jacobi polynomials, and for which we also used the notation $\jac_n^{(a, b)}$.

@@

Further, note that $J_x^{(a,b,c,d)}, J_y^{(a,b,c,d)}$ are banded-block-banded matrices (see \bsrefdef{def:bandedblockbanded}). They are block-tridiagonal (block-bandwidths $(1,1)$):
\bseqn{
	J_{x/y}^{(a,b,c,d)} &= 
		\begin{pmatrix}
			B^{x/y}_0 & A^{x/y}_0 & & & & \\
			C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
			& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
			& & C^{x/y}_3 & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}.
}
For $J_x^{(a,b,c,d)}$, the blocks are diagonal (sub-block-bandwidths $(0,0)$) for $n = 0,1,2,\dots$:
\bseqn{
	A^x_n &:= 
		\begin{pmatrix}
			\alphaabcd_{n+1,0,1} & 0 & \hdots & 0 \\
			& \ddots & & \vdots & \\
			& & \alphaabcd_{n+1,n,1} & 0 \\
		\end{pmatrix} \in \R^{(n+1)\times(n+2)}, \\
	B^x_n &:= 
		\begin{pmatrix}
			\alphaabcd_{n,0,2} & & \\
			& \ddots & \\
			& & \alphaabcd_{n,n,2} \\
		\end{pmatrix} \in \R^{(n+1)\times(n+1)}, \\
	C^x_n &:= \big( A^x_{n-1} \big)^\top \in \R^{(n+1)\times n}, \quad (n \ne 0)
}
and for $J_y^{(a,b,c,d)}$ the blocks are tridiagonal (sub-block-bandwidths $(1,1)$):
\bseqn{
	A^y_n &:= 
		\begin{pmatrix}
			\betaabcd_{n,0,8} & \betaabcd_{n,0,9} & & & \\
			\betaabcd_{n,1,7}& \ddots & \ddots & & \\
			& \ddots & \ddots & \ddots & \\
			& & \betaabcd_{n,n,7} & \betaabcd_{n,n,8} & \betaabcd_{n,n,9} \\
		\end{pmatrix} \in \R^{(n+1)\times(n+2)}, \\
	B^y_n &:= 
		\begin{pmatrix}
			\betaabcd_{n,0,5} & \betaabcd_{n,0,6} & & \\
			\betaabcd_{n,1,4} & \ddots & \ddots & \\
			& \ddots & \ddots & \betaabcd_{n,n-1,6} \\
			& & \betaabcd_{n,n,4} & \betaabcd_{n,n,5}
		\end{pmatrix} \in \R^{(n+1)\times(n+1)}, \\
	C^y_n &:= 
		\begin{pmatrix}
			\betaabcd_{n,0,2} & \betaabcd_{n,0,3} & & \\
			\betaabcd_{n,1,1} & \ddots & \ddots & \\
			& \ddots & \ddots & \betaabcd_{n,n-2,3} \\
			& & \ddots & \betaabcd_{n,n-1,2} \\
			& & & \betaabcd_{n,n,1} \\
		\end{pmatrix} \in \R^{(n+1)\times n}, \quad (n \ne 0).
}
Note that the sparsity of the Jacobi matrices (in particular the sparsity of the sub-blocks) comes from the natural sparsity of the three-term recurrences of the 1D OPs, meaning that the sparsity is not limited to the specific disk-slice case.

@@

In this section, we concentrate on the disk-slice case, and simply note that similar arguments apply for the trapezium case. Recall that, for the disk-slice,
\bseqn{
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
}
where
\bseqn{
	\begin{cases}
		(\alpha,\beta) &\subset (0,1) \\
		(\gamma,\delta) &:= (-1,1) \\
		\rho(x) &:= (1-x^2)^{\half}
	\end{cases}.
}
The 2D OPs on the disk-slice $\Omega$, orthogonal with respect to the weight
\bseqn{
	\Wabc(x,y) &:= \genjacw^{(a,b,2c)}(x) \: \jacw^{(c)}\fpr(\frac{y}{\rho(x)}) \\
	&= (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c, \quad (x,y) \in \Omega,
}
are then given by:
\bseqn{
	\hdopnkabc(x,y) := \genjacnmk^{(a, b, 2c+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(c,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega
}
where the 1D OPs $\{\genjac^{(a, b, c)}_n\}$ are orthonormal on the interval $(\alpha, \beta)$ with respect to the weight
\bseqn{
	\genjacw^{(a,b,c)}(x) :=  (\beta - x)^a \: (x - \alpha)^b \: \rho(x)^{c}
}
and the 1D OPs $\{\jac^{(c, c)}_n\}$ are orthonormal on the interval $(\gamma, \delta) = (-1, 1)$ with respect to the weight
\bseqn{
	\jacw^{(c)}(x) := (1-x)^c \: (1+x)^c = (1-x^2)^c.
}
Denote the weighted OPs by
\bseqn{
	\bigWabc(x,y) := \Wabc(x,y) \: \bighdopabc(x,y),
}
and recall that a function $f(x,y)$ defined on $\Omega$ is approximated by its expansion $f(x,y) = \bighdopabc(x,y)^\top \vec{f}$.

@@

The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomial families' derivatives, for example the Jacobi polynomials on the interval, as seen in the DLMF \cite[(18.9.3)]{DLMF}, and on the triangle \cite{olver2018recurrence}. An illustration of how the non-weighted differential operators increment the parameters $(a,b,c)$ is seen in \bsreffig{fig:ds:laplaceparameterconversion}.
